\bf Interactive Visualization of Hierarchically Structured Data

### 2017-03-23 treelapse.tex

---
standalone: true
documentclass: article
fontsize: 12pt
output:
  bookdown::pdf_book:
    keep_tex: yes
    toc: false
    includes:
      in_header: preamble.tex
      before_body: before_body.tex
bibliography: ["bibliography.bib"]
biblio-style: "agsm"
link-citations: yes
---

## Introduction ##

```{r set_chunk_opts, echo = FALSE}
library("knitr")
opts_chunk$set(
  fig.align = "center",
  out.width = "350px",
  echo = FALSE,
  cache = TRUE
)
```

We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series. We
hope both to characterize generically useful techniques for
interactively visualizing hierarchical data and to offer practical
tools for implementing such displays. To this end, we identify
questions that often emerge when working with hierarchical data and
provide an R package to simplify their investigation [@ihaka1996r].

Our key contribution is the adaptation of the visualization principles
of focus-plus-context and linking to the study of tree-structured data
[@buja1996interactive, @becker1987brushing]. Our motivating
application is to the analysis of bacterial time series, where an
evolutionary tree relating bacteria is available a priori. However, we
have identified common problem types where, if a tree is not directly
available, it can be constructed from data and then studied using our
techniques.

@@

### Problem Motivation ###

A microbiome is a community of bacteria living in given environments,
for example, ocean water or the human gut [@human2012structure,
@human2012structure, @cho2012human]. Progress in the field has been
rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of bacterial ecological structure and
its influence in human and environmental health. Being concerned with
both bacterial community structure and human health, the field exists
at the border between ecology and medicine; consequently, papers in
the area often apply a blend of exploratory data analysis and formal
statistical inference.

The two essential microbiome analysis problems that motivated our work
are the tree-structured differential abundance and differential
dynamics problems. In the differential abundance problem, we attempt
to compare the abundances of individual bacteria across experimental
conditions -- for example, treatment vs. control or healthy
vs. diseased. This is the microbiome analog of differential expression
analysis in genomics [@anders2010differential]. We prepend the
description "tree-structured" because, in practice, researchers generate
interpretations about intermediate taxonomic orders -- it is more interesting to
discover novel behavior taxonomic levels between high-order phyla and low-level
species. Hence, we frame the tree-structured differential abundance problem as
the question of identifying the largest taxonomic subtree whose associated
bacteria are differentially abundant.

@@

### Problem Abstraction ###

To unify the tree-structured differential abundance and bacterial
dynamics problems, we identify the data with a collection of random
variables indexed by nodes in a prespecified tree structure. In the
differential abundance problem, each random variable lives in
$\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the abundance for that group, and
a node exhibits differential abundance when the coordinates are drawn
from different distributions. On the other hand, in the bacterial
dynamics problem, each random variable is a time series, living in
$\mathbb{R}^{T}$.

@@

This framework suggests other potential application areas, not all of
which have a priori known tree structures. For example, collections of
spatially-indexed time series are frequently encountered in practice
-- consider energy consumption, product sales, or high school dropout
rates across regional districts. This type of data has an implicit
tree structure -- at the top level are different states, while at the
bottom are individual census tracts, say. Analysis here revolves
around the question of how variation across time series is related to
their geographic position.  A case study to this type of data is given
in Section \@ref(zillow-study).

Alternatively, if this type of hierarchical contextual information is
not directly available, a tree structure can be learned from the
data. This could be achieved by learning a hierarchical clustering on
the original series. Further, if contextual information is available,
but it is not hierarchical, it is possible to setup a supervised
problem that uses context to predict features of the time series. We
can construct a tree by applying a tree-based classifier
[@breiman1984classification] or extracting a regression tree from a
more complex supervised model [@boz2002extracting,
@saito2002extracting]. Analysis then focuses on how different
partitions of the contextual, covariate space relate to observed time
series. This approach is described in Section \@ref(bikesharing-study).

Finally, note that, while we have focused on time series valued nodes,
all of this discussion could be translated to studying
high-dimensional data via parallel coordinates
[@inselberg1991parallel]. The usual parallel coordinates challenges
remain, mainly selecting scales for and ordering across the different
coordinates, but the linking and focus-plus-context can still be
employed this setting.

## Background Literature and Solution Principles ##

@@

The focus-plus-context principle is that large collections of visual
elements can be studied at multiple scales, by simultaneously
focusing" on a few elements of interest and maintaining the "context"
of the coarser-scale background. A simple example of this idea is to
include a search box that highlights matching samples (focus) and
mutes the rest (context). Two more sophisticated methods anchored in
this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
central to the proposals in treelapse [@hochheiser2004dynamic,
@heer2004doitrees]. In timeboxes, a collection of time series are
graphically queried using interactive brushes. Series that pass
through all of the user-specified brushes are highlighted, and the
rest are faded to the background. Hence, time series meeting the
constraints imposed by the brushes are focused, while the remainder
are de-emphasized, though they remain present as context. This method
can be interpreted programmatically as the visual analog of a database
query, or probabilistically as the conditional distribution for the
full series, given it passes through certain bounds.

In DOI trees, the viewer's attention is focused on a collection of
high-interest nodes, while the remaining lower-interest nodes are left
on the fringes as context. The implementation is modularized into two
tasks -- the determination of a DOI distribution over nodes in the
tree and visual layout of a tree given DOI assignments. The DOI
distribution used in [@heer2004doitrees] places maximal interest on
the node that the user had most recently clicked, along with all
ancestors. The DOI for all other nodes is defined as the distance to
the closest maximal interest node. The layout step then trims
low-interest subtrees until the remaining nodes fit within a given
screen size. By adjusting the minimal DOI below which nodes are
hidden, the user can transition between node-specific and full-tree
scales.

In linking, alternative representations of the same samples are placed
side-by-side in order to display covariation across views. A canonical
application is to linked scatterplot brushing
[@becker1987brushing]. Here, a scatterplot matrix gives the
relationship between all pairs of variables.  Points brushed in one
scatterplot are then highlighted in all others. For example, this
helps the user determine whether an outlier in one dimension is an
outlier in others. Another instance of this idea links the results of
dimensionality reduction methods to displays of the raw data, as
implemented by XGobi and Cranvas, for example [@xie2013cranvas,
@swayne1998xgobi]. As in timeboxes, linking can be interpreted as
database queries or conditional probabilities: given a subset of the
series after conditioning on the values for one set of features, what
are the values for a second set [@buja1996interactive]?

Finally, unrelated to established visualization principles, we note
that our work is deliberately grounded in the R software
ecosystem. This connection is made using the htmlwidgets package
[@vaidyanathan2014htmlwidgets]. Not only does linking R with D3 make
these visualization methods more broadly accessible, we hope to
facilitate exchange between data modeling and interactive
visualization. Moreover, our tools are intentionally limited in scope
-- designed to facilitate this dialog for a specific class of
problems, rather than providing a toolbox for generic types of
visualization design. We believe that this narrow context within a
broad ecosystem strikes a balance between problem-specificity and
ease-of-use.

## Specific Proposals ##

Our first proposed visualization technique is a minor modification of
the DOI tree. The standard DOI tree definition does not have any
notion of data defined at nodes, it is only used a device for
navigating tree structures. A trivial extension can encode scalar data
at nodes: have the node radius reflect the associated scalar value. To
reinforce this effect, we can adjust the width of the parent
edge. When parent nodes have values equal to the sum of their
children, this creates the effect of values "flowing" from the root to
leaves. To help viewers make use of their domain knowledge, we have
included a search box that highlights paths to nodes with matching
terms. Edges are ordered from widest on the left to narrowest on the
right. While this method can only represent a single scalar-value per
tree node, it suggests an approach to the tree-structured differential
abundance problem, which we call the DOI sankey.

In the DOI sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment
and control groups at each tree node. Every edge in the tree is split
into two colors^[We use the colorbrewer palette to facilitate
readability [@brewer2003colorbrewer]], with relative widths of the
different colors reflecting differences in sizes for the two
groups. The overall width of each edge represents the sum of values
across all groups.

This display is designed to facilitate investigation of the tree
structured differential abundance question. For example, for a single
node and a single group, first compute the average abundance at that
node among all samples in that group. This will give the width for
that group's color on the edge leading to the specified
node. Differentially abundant subtrees then correspond to subtrees
where some colors occupy more space than others. That is, this
representation makes it easier to identify points where the "flows"
for different groups diverge -- the colors begin to separate. The DOI
principle assists navigation across the tree structure, allowing focus
on individual flow structures without losing broader tree context.

Our third display is directed at the bacterial dynamics
question. Here, two panels are arranged one over the other; one
displays all time series, while the other displays all tree nodes,
with node sizes reflecting the time series value at that node. For
this reason, we call the display, Timebox Trees. In the time series
panel, we have directly implemented the timeboxes idea. We then link
the panels: when a set of series is highlighted by the timeboxes, the
associated tree nodes are also highlighted. For example, timeboxes can
be used to focus on a set of series with specific shape -- increased
abundance after an ecological shock, for example -- and identify along
what subtrees this pattern is present. To further focus on specific
elements, a pan-zoom scented widget is provided
[@willett2007scented]. The widget is a miniature version of the full
time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and
sankeys, a search bar can be used to highlight those series of
interest a priori.

@@

## Case Studies ##

@@

### Bacterial Dynamics of Antibiotics Time Courses ###

@dethlefsen2008pervasive investigated the effect of antibiotics on
bacterial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across bacteria is important, considering
the the role of bacteria in health and not just disease.

We approach the data using the linked time and treebox views, after
first filtering low variance taxa and taking an $\text{asinh}$
transformation. An initial view, Figure \@ref(fig:antibioticoverview),
reveals two dramatic drops in the overall bacterial abundance time
series during the antibiotics time courses. Two more subtle effects
are also suggested from this view,

* The second antibiotic treatment seems to have a more lasting effect, as the
series take longer to return to their original values.
* Some high level taxa appear relatively unaffected by the first
antibiotic treatment.  By more closely inspecting the display, we are
able to identify these as members of the Bacteroidetes phylum, see
Figure \@ref(fig:antibioticbacteroidetes).

```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa. We have annotated the figure with its three main components."}
include_graphics("figure/annotated_antibiotic_overview.png")
```

```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
include_graphics("figure/antibiotic_bacteroidetes.png")
```

Next, using the scented widget, we focus on the window around the
second antibiotic treatment. We apply the treebox display to compare
then behavior of different families of Firmicutes, Lachnospiraceae and
Ruminoccocus. We suspect that these taxa are associated with the
delayed recovery after the second time course. To investigate this, we
input these family names in the search box to isolate their positions
on the tree; then we apply brushes to highlight the series that
contribute to these higher-level families. The resulting view is given
by Figure \@ref(fig:antibioticfirmicutes)

```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
include_graphics("figure/antibiotic_firmicutes.png")
```

Alternatively, we can summarize each node by the average across its
descendants -- this brings attention to individual bacteria that may
be underlying some of the broader taxonomic patterns we have noted
when studying the subtree sums. For example, in Figure
\@ref(fig:ruminococcus), we highlight all families below order
Ruminococcus, suggesting that the decrease due to antibiotics occurs
uniformly across almost all families. A point that was not evident in
the earlier sum-across-descendants view is that, after the second
treatment of antibiotics, a few of the Ruminoccocus families recover
more rapidly than the rest, for example the Unc095d3 (highlighted in
red) are only briefly affected. In contrast, most families seem to
recover in unison after the first treatment.

```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders."}
include_graphics("figure/ruminococcus.png")
```

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated --
because there are not many leaves associated with this phylum, the sum
was small. Interestingly, the abundance of these bacteria seems to
*increase* after the first antibiotics treatment.  Be cautious,
however, that the average over only a few Verrucomicrobiae species
will be a more variable estimate than the averages over the more
prevalent phyla.

```{r verrucomicrobiae, fig.cap = "The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant. In particular, they seem to increase after the first antibiotic time course."}
include_graphics("figure/verrucomicrobiae.png")
```

### Differential Bacterial Abundance and Preterm Births ###

@digiulio2015temporal tracked the abundance of bacteria in
the vaginal microbiome during pregnancy in an effort to study
relationships between bacterial community composition and preterm
birth. Ideally, it would be possible to develop clear bacterial
signatures associated with preterm births.

Unlike the antibiotics study, we have measurements across more
individuals than we could reasonably inspect one at a time. While we
could average across all individuals, we will take our cue from
[@digiulio2015temporal] and place each sample into one of five
Community State Types (CSTs), identified via k-medoids.  In that
study, a linear model identified one of these CSTs (CST 4) as
significantly more diverse, further it appeared associated with
preterm births. Here, we corroborate this finding using exploratory
views.

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
a heatmap, with bacteria ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual bacteria loses power -- we can compare
bacterial abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (species-level)
leaf nodes, and then aggregated the averages up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundance of these bacteria distinguishes the
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

Alternatively, in Figure \@ref(fig:pretermcsts), we avoid working with
CSTs, displaying instead averages among samples associated with either
preterm or term births. The green edges are associated with preterm
births -- we see that they contribute more weight to phyla outside the
Firmicutes. This is consistent with the claim that CST 4, the most
diverse of the CSTs, is associated with preterm births.

```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
include_graphics("figure/preterm_csts.png")
```

```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
include_graphics("figure/preterm_csts_lacto.png")
```

```{r pretermpreterm, out.width = "300px", fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
include_graphics("figure/preterm_preterm.png")
```

### Dynamics in Housing Prices {#zillow-study}

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See Section
\@ref(conclusion) for potential optimizations, however.].

>>>>

\documentclass[12pt]{article}
\usepackage{hyperref}

\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Interactive Visualization of Hierarchically Structured Data}
  \author{Kris Sankaran\thanks{
      This research was supported in part by an NIH training grant (5T32GM096982-03).}\hspace{.2cm}\\
    Department of Statistics, Stanford University\\
    and \\
    Susan Holmes\thanks{
       This research was supported in part by NIH grant R01 AI112401.
      }\hspace{0.2cm}\\
    Department of Statistics, Stanford University}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series.  To
this end, we identify questions that often emerge when working with
hierarchical data and provide an R package to simplify their
investigation. Our key contribution is the adaptation of the
visualization principles of focus-plus-context and linking to the
study of tree-structured data.

Our motivating application is to the analysis of microbial time
series, where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques. We perform detailed case studies to
describe the alternative use cases, interpretations, and utility of
the proposed visualization methods.

\end{abstract}

\noindent%
{\it Keywords:}  D3, focus-plus-context, linking, time-series, tree-structured, R
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\subsection*{Introduction}\label{introduction}

We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series. We hope
both to characterize generically useful techniques for interactively
visualizing hierarchical data and to offer practical tools for
implementing such displays. To this end, we identify questions that
often emerge when working with hierarchical data and provide an R
package to simplify their investigation \citep{ihaka1996r}.

Our key contribution is the adaptation of the visualization principles
of focus-plus-context and linking to the study of tree-structured data
\citep{buja1996interactive, becker1987brushing}. Our motivating
application is to the analysis of bacterial time series, where an
evolutionary tree relating bacteria is available a priori. However, we
have identified common problem types where, if a tree is not directly
available, it can be constructed from data and then studied using our
techniques.

@@

\subsubsection*{Problem Motivation}\label{problem-motivation}

A microbiome is a community of bacteria living in given environments,
for example, ocean water or the human gut 
\citep{human2012structure, cho2012human}. Progress in the field has
been rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of bacterial ecological structure and its
influence in human and environmental health. Being concerned with both
bacterial community structure and human health, the field exists at the
border between ecology and medicine; consequently, papers in the area
often apply a blend of exploratory data analysis and formal statistical
inference.

The two essential microbiome analysis problems that motivated our work
are the tree-structured differential abundance and differential dynamics
problems. In the differential abundance problem, we attempt to compare
the abundances of individual bacteria across experimental conditions --
for example, treatment vs. control or healthy vs. diseased. This is the
microbiome analog of differential expression analysis in genomics
\citep{anders2010differential}.
We prepend the description ``tree-structured'' because, in practice,
researchers generate interpretations about intermediate taxonomic orders
-- it is more interesting to discover novel behavior taxonomic levels
between high-order phyla and low-level species. Hence, we frame the
tree-structured differential abundance problem as the question of
identifying the largest taxonomic subtree whose associated bacteria are
differentially abundant.

@@

\subsubsection*{Problem Abstraction}\label{problem-abstraction}

To unify the tree-structured differential abundance and bacterial
dynamics problems, we identify the data with a collection of random
variables indexed by nodes in a prespecified tree structure. In the
differential abundance problem, each random variable lives in
\(\mathbb{R}^{G}\) where \(G\) is the number of groups being compared.
Each coordinate represents the abundance for that group, and a node
exhibits differential abundance when the coordinates are drawn from
different distributions. On the other hand, in the bacterial dynamics
problem, each random variable is a time series, living in
\(\mathbb{R}^{T}\).

@@

This framework suggests other potential application areas, not all of
which have a priori known tree structures. For example, collections of
spatially-indexed time series are frequently encountered in practice --
consider energy consumption, product sales, or high school dropout rates
across regional districts. This type of data has an implicit tree
structure -- at the top level are different states, while at the bottom
are individual census tracts, say. Analysis here revolves around the
question of how variation across time series is related to their
geographic position. A case study to this type of data is given in
Section \ref{zillow-study}.

Alternatively, if this type of hierarchical contextual information is
not directly available, a tree structure can be learned from the data.
This could be achieved by learning a hierarchical clustering on the
original series. Further, if contextual information is available, but it
is not hierarchical, it is possible to setup a supervised problem that
uses context to predict features of the time series. We can construct a tree by
applying a tree-based classifier \citep{breiman1984classification} or extracting
a regression tree from a more complex supervised model
\citep{boz2002extracting,saito2002extracting}. Analysis then focuses on how
different partitions of the contextual, covariate space relate to observed time
series. This approach is described in Section \ref{bikesharing-study}.

Finally, note that, while we have focused on time series valued nodes,
all of this discussion could be translated to studying high-dimensional
data via parallel coordinates \citep{inselberg1991parallel}. The usual parallel
coordinates challenges remain, mainly selecting scales for and ordering
across the different coordinates, but the linking and focus-plus-context
can still be employed this setting.

\subsection*{Background Literature and Solution
Principles}\label{background-literature-and-solution-principles}

@@

The focus-plus-context principle is that large collections of visual
elements can be studied at multiple scales, by simultaneously focusing"
on a few elements of interest and maintaining the ``context'' of the
coarser-scale background. A simple example of this idea is to include a
search box that highlights matching samples (focus) and mutes the rest
(context). Two more sophisticated methods anchored in this idea are
timeboxes and Degree-of-Interest (DOI) trees; both are central to the proposals
in treelapse \citep{hochheiser2004dynamic, heer2004doitrees}. In timeboxes, a
collection of time series are graphically queried using interactive brushes.
Series that pass through all of the user-specified brushes are highlighted, and
the rest are faded to the background. Hence, time series meeting the constraints
imposed by the brushes are focused, while the remainder are de-emphasized,
though they remain present as context. This method can be interpreted
programmatically as the visual analog of a database query, or probabilistically
as the conditional distribution for the full series, given it passes through
certain bounds.

In DOI trees, the viewer's attention is focused on a collection of
high-interest nodes, while the remaining lower-interest nodes are left
on the fringes as context. The implementation is modularized into two
tasks -- the determination of a DOI distribution over nodes in the tree
and visual layout of a tree given DOI assignments. The DOI distribution
used in \citep{heer2004doitrees}
places maximal interest on the node that the user had most recently
clicked, along with all ancestors. The DOI for all other nodes is
defined as the distance to the closest maximal interest node. The layout
step then trims low-interest subtrees until the remaining nodes fit
within a given screen size. By adjusting the minimal DOI below which
nodes are hidden, the user can transition between node-specific and
full-tree scales.

In linking, alternative representations of the same samples are placed
side-by-side in order to display covariation across views. A canonical
application is to linked scatterplot brushing 
\citep{becker1987brushing}. Here, a scatterplot
matrix gives the relationship between all pairs of variables. ~Points
brushed in one scatterplot are then highlighted in all others. For
example, this helps the user determine whether an outlier in one
dimension is an outlier in others. Another instance of this idea links
the results of dimensionality reduction methods to displays of the raw
data, as implemented by XGobi and Cranvas, for example 
\citep{xie2013cranvas, swayne1998xgobi}. As in timeboxes,
linking can be interpreted as database queries or conditional
probabilities: given a subset of the series after conditioning on the
values for one set of features, what are the values for a second set
\citep{buja1996interactive}?

Finally, unrelated to established visualization principles, we note that
our work is deliberately grounded in the R software ecosystem. This
connection is made using the htmlwidgets package 
\citep{vaidyanathan2014htmlwidgets}. Not only
does linking R with D3 make these visualization methods more broadly
accessible, we hope to facilitate exchange between data modeling and
interactive visualization. Moreover, our tools are intentionally limited
in scope -- designed to facilitate this dialog for a specific class of
problems, rather than providing a toolbox for generic types of
visualization design. We believe that this narrow context within a broad
ecosystem strikes a balance between problem-specificity and ease-of-use.

\subsection*{Specific Proposals}\label{specific-proposals}

Our first proposed visualization technique is a minor modification of
the DOI tree. The standard DOI tree definition does not have any notion
of data defined at nodes, it is only used a device for navigating tree
structures. A trivial extension can encode scalar data at nodes: have
the node radius reflect the associated scalar value. To reinforce this
effect, we can adjust the width of the parent edge. When parent nodes
have values equal to the sum of their children, this creates the effect
of values ``flowing'' from the root to leaves. To help viewers make use
of their domain knowledge, we have included a search box that highlights
paths to nodes with matching terms. Edges are ordered from widest on the
left to narrowest on the right. While this method can only represent a
single scalar-value per tree node, it suggests an approach to the
tree-structured differential abundance problem, which we call the DOI
sankey.

In the DOI sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment
and control groups at each tree node. Every edge in the tree is split
into two colors\footnote{We use the colorbrewer palette to facilitate
  readability \citep{brewer2003colorbrewer}}, with relative
widths of the different colors reflecting differences in sizes for the
two groups. The overall width of each edge represents the sum of values
across all groups.

This display is designed to facilitate investigation of the tree
structured differential abundance question. For example, for a single
node and a single group, first compute the average abundance at that
node among all samples in that group. This will give the width for that
group's color on the edge leading to the specified node. Differentially
abundant subtrees then correspond to subtrees where some colors occupy
more space than others. That is, this representation makes it easier to
identify points where the ``flows'' for different groups diverge -- the
colors begin to separate. The DOI principle assists navigation across
the tree structure, allowing focus on individual flow structures without
losing broader tree context.

Our third display is directed at the bacterial dynamics question. Here,
two panels are arranged one over the other; one displays all time
series, while the other displays all tree nodes, with node sizes
reflecting the time series value at that node. For this reason, we call
the display, Timebox Trees. In the time series panel, we have directly
implemented the timeboxes idea. We then link the panels: when a set of
series is highlighted by the timeboxes, the associated tree nodes are
also highlighted. For example, timeboxes can be used to focus on a set
of series with specific shape -- increased abundance after an ecological
shock, for example -- and identify along what subtrees this pattern is
present. To further focus on specific elements, a pan-zoom scented
widget is provided \citep{willett2007scented}. The widget is a miniature version
of the full time series panel, equipped with a single brush whose extent
specifies the limits in the main time series panel. As in the DOI trees and
sankeys, a search bar can be used to highlight those series of interest a
priori.

@@

\subsection*{Case Studies}\label{case-studies}

@@

\subsubsection*{Bacterial Dynamics of Antibiotics Time
Courses}\label{bacterial-dynamics-of-antibiotics-time-courses}

\citet{dethlefsen2008pervasive} investigated the
effect of antibiotics on bacterial community composition from an
ecological perspective. The study tracks the microbiome of three
patients across ten months, with two five-day antibiotic time courses
separated by 6 months. Discerning the variation in resilience across
bacteria is important, considering the the role of bacteria in health
and not just disease.

We approach the data using the linked time and treebox views, after
first filtering low variance taxa and taking an \(\text{asinh}\)
transformation. An initial view, Figure \ref{fig:antibioticoverview},
reveals two dramatic drops in the overall bacterial abundance time
series during the antibiotics time courses. Two more subtle effects are
also suggested from this view,

\begin{itemize}
\item
  The second antibiotic treatment seems to have a more lasting effect,
  as the series take longer to return to their original values.
\item
  Some high level taxa appear relatively unaffected by the first
  antibiotic treatment. By more closely inspecting the display, we are
  able to identify these as members of the Bacteroidetes phylum, see
  Figure \ref{fig:antibioticbacteroidetes}.
\end{itemize}

\begin{figure}

{\centering \includegraphics[width=350px]{figure/annotated_antibiotic_overview}

}

\caption{The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa. We have annotated the figure with its three main components.}\label{fig:antibioticoverview}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=350px]{figure/antibiotic_bacteroidetes}

}

\caption{Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics.}\label{fig:antibioticbacteroidetes}
\end{figure}

Next, using the scented widget, we focus on the window around the second
antibiotic treatment. We apply the treebox display to compare then
behavior of different families of Firmicutes, Lachnospiraceae and
Ruminoccocus. We suspect that these taxa are associated with the delayed
recovery after the second time course. To investigate this, we input
these family names in the search box to isolate their positions on the
tree; then we apply brushes to highlight the series that contribute to
these higher-level families. The resulting view is given by Figure
\ref{fig:antibioticfirmicutes}

\begin{figure}

{\centering \includegraphics[width=350px]{figure/antibiotic_firmicutes}

}

\caption{Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
\end{figure}

Alternatively, we can summarize each node by the average across its
descendants -- this brings attention to individual bacteria that may be
underlying some of the broader taxonomic patterns we have noted when
studying the subtree sums. For example, in Figure
\ref{fig:ruminococcus}, we highlight all families below order
Ruminococcus, suggesting that the decrease due to antibiotics occurs
uniformly across almost all families. A point that was not evident in
the earlier sum-across-descendants view is that, after the second
treatment of antibiotics, a few of the Ruminoccocus families recover
more rapidly than the rest, for example the Unc095d3 (highlighted in
red) are only briefly affected. In contrast, most families seem to
recover in unison after the first treatment.

\begin{figure}

{\centering \includegraphics[width=350px]{figure/ruminococcus}

}

\caption{By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders.}\label{fig:ruminococcus}
\end{figure}

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the Verrucomicrobiae
at the far left, which seem to have large abundance over many time
points. This phylum had been previously obfuscated -- because there are
not many leaves associated with this phylum, the sum was small.
Interestingly, the abundance of these bacteria seems to \emph{increase}
after the first antibiotics treatment. Be cautious, however, that the
average over only a few Verrucomicrobiae species will be a more variable
estimate than the averages over the more prevalent phyla.

\begin{figure}

{\centering \includegraphics[width=350px]{figure/verrucomicrobiae}

}

\caption{The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant. In particular, they seem to increase after the first antibiotic time course.}\label{fig:verrucomicrobiae}
\end{figure}

\subsubsection*{Differential Bacterial Abundance and Preterm
Births}\label{differential-bacterial-abundance-and-preterm-births}

\citet{digiulio2015temporal}
tracked the abundance of bacteria in the vaginal microbiome during
pregnancy in an effort to study relationships between bacterial
community composition and preterm birth. Ideally, it would be possible
to develop clear bacterial signatures associated with preterm births.

Unlike the antibiotics study, we have measurements across more
individuals than we could reasonably inspect one at a time. While we
could average across all individuals, we will take our cue from
\citep{digiulio2015temporal}
place each sample into one of five Community State Types (CSTs),
identified via k-medoids. In that study, a linear model identified one
of these CSTs (CST 4) as significantly more diverse, further it appeared
associated with preterm births. Here, we corroborate this finding using
exploratory views.

Therefore, our focus here is on the differential abundance question,
rather than dynamics. We would like to provide visual representations of
differential abundance across CSTs and also between preterm and non-preterm
births. \citet{digiulio2015temporal} interpreted the CSTs using a heatmap, with
bacteria ordered according to a hierarchical clustering. By using the DOI sankey
instead, we can interpret the CSTs in their taxonomic context and at multiple
scales of taxonomic resolution. Further, while 
\citep{digiulio2015temporal} focused on identifying associations between
preterm births and CSTs -- presumably because testing individual bacteria loses
power -- we can compare bacterial abundances between preterm and non-preterm
samples along subtrees, without requiring CSTs as an intermediary.

In Figure \ref{fig:pretermcsts}, we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of all
samples within each CST to define values at the (species-level) leaf
nodes, and then aggregated the averages up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundance of these bacteria distinguishes the
remaining CSTs, see Figure \ref{fig:pretermcstslacto}.

Alternatively, in Figure \ref{fig:pretermcsts}, we avoid working with
CSTs, displaying instead averages among samples associated with either
preterm or term births. The green edges are associated with preterm
births -- we see that they contribute more weight to phyla outside the
Firmicutes. This is consistent with the claim that CST 4, the most
diverse of the CSTs, is associated with preterm births.

\begin{figure}

{\centering \includegraphics[width=350px]{figure/preterm_csts}

}

\caption{The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes.}\label{fig:pretermcsts}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=350px]{figure/preterm_csts_lacto}

}

\caption{Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent.}\label{fig:pretermcstslacto}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=300px]{figure/preterm_preterm}

}

\caption{Samples with high levels of phyla other than Firmicutes appear to be related to preterm births.}\label{fig:pretermpreterm}
\end{figure}

\subsubsection*{Dynamics in Housing Prices}\label{zillow-study}
